---
title: "Get started with dlmoR"
description: "Estimate dim-light melatonin onset from a melatonin concentration profile"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)

article_figure <- function(path) {
  source <- system.file(path, package = "dlmoR")
  if (identical(source, "")) {
    stop("Could not find package figure: ", path, call. = FALSE)
  }
  dir.create("figures", showWarnings = FALSE)
  target <- file.path("figures", basename(path))
  file.copy(source, target, overwrite = TRUE)
  target
}
```

`dlmoR` estimates dim-light melatonin onset (DLMO) from time series of
melatonin concentrations using the hockey-stick method. The main function is
`calculate_dlmo()`, which labels the profile, defines the fitting region, runs a
grid search, and returns DLMO estimates together with diagnostic plots.

This vignette will guide you in not only obtaining a DLMO timestamp, but in
understanding how the analysis settings affect the labelled profile, the
selected rise, and the diagnostic plots.

## Installation

Install the development version of the `dlmoR` package from GitHub:

```{r}
#| eval: false
# install.packages("pak")
pak::pak("tscnlab/dlmoR")
```

Then load the package:

```{r}
library(dlmoR)
```

## Example profiles

`calculate_dlmo()` expects a data frame, tibble, or CSV file with two columns:

- `datetime`: sampling time as `POSIXct`, or a value that can be converted to
  `POSIXct`
- `melatonin`: numeric melatonin concentration, in pg/mL

This vignette uses example profiles included with the package.

## Plot a raw profile

Before fitting a model, inspect the raw melatonin time series. A raw plot gives
you a first look at the shape of the profile before fitting the hockey-stick
model, including possible threshold crossings, isolated high points, or multiple
rises.

```{r}
profile_a <- system.file(
  "extdata/vignette_profiles/melatonin_profile_a.csv",
  package = "dlmoR"
)
```

```{r}
#| label: fig-raw-no-threshold
#| fig-cap: "*Raw melatonin profile without a threshold overlay.*"
plot_raw_profile(file_path = profile_a)
```

Adding a threshold line can make the first visual check more informative:

```{r}
#| label: fig-raw-with-threshold
#| fig-cap: "*Raw melatonin profile with a threshold overlay.*"
plot_raw_profile(file_path = profile_a, threshold = 5)
```

You can also read the data yourself and use `plot_profile(..., mode = "raw")`:

```{r}
melatonin_profile <- readr::read_delim(profile_a, delim = ";", show_col_types = FALSE)
plot_profile(melatonin_profile, threshold = 5, mode = "raw")
```

## Basic DLMO estimate

For exploratory work, it is often useful to start with the coarse grid search
only. This is faster and lets you check whether the profile segmentation and
region of interest look sensible.

```{r}
dlmo_coarse <- calculate_dlmo(
  file_path = profile_a,
  threshold = 5,
  fine_flag = FALSE
)
```

The coarse result is returned in `dlmo_coarse$dlmo$coarse`. This includes the
estimated DLMO time, the fitted melatonin concentration at that time, and the
parameters for the fitted hockey-stick lines:

```{r}
dlmo_coarse$dlmo$coarse
```

The threshold and interval limit used for the calculation are also stored in the
result object:

```{r}
dlmo_coarse$threshold
dlmo_coarse$interval_limit
```

The labelled melatonin profile is stored in `prof`. These labels are the bridge
between the raw measurements and the fit shown in the diagnostic plot:

```{r}
dlmo_coarse$prof
```

The most important segment labels are:

- `base`: baseline points used for the early part of the hockey-stick fit
- `ascending`: points used for the melatonin rise
- `intermediate`: points between base and ascending segments, when present

The coarse diagnostic plot is stored in `dlmoplotcoarse`:

```{r}
#| label: fig-coarse-diagnostic
#| fig-cap: "*Coarse DLMO diagnostic plot. Blue points are the base segment, green points are the ascending segment, the horizontal line is the threshold, and the diamond marks the estimated DLMO.*"
dlmo_coarse$dlmoplotcoarse
```

Look for three things:

- the base segment should represent the relatively stable pre-rise portion of
  the profile
- the ascending segment should follow the biologically plausible melatonin rise
- the DLMO marker should fall near the transition between baseline and rise,
  rather than on an isolated noisy point

For a more accurate estimate, run the fine search:

```{r}
dlmo_result <- calculate_dlmo(
  file_path = profile_a,
  threshold = 5,
  fine_flag = TRUE
)

dlmo_result$dlmo$fine$time
dlmo_result$dlmoplotfine
```

The fine search is more computationally expensive because it refines the grid
around the best coarse-grid solutions. For profiles with a wide region of
interest, this can take noticeably longer.

## Output structure

`calculate_dlmo()` returns a list with:

- `prof`: segmented melatonin profile
- `prl`: parallelogram diagnostic information
- `roi`: region of interest for the grid search
- `threshold`: threshold used for the calculation
- `interval_limit`: interval-limit setting used for repeated crossing selection
- `ip`: inflection-point search details
- `dlmo`: coarse and, if requested, fine DLMO summaries
- `dlmoplotcoarse`: coarse diagnostic plot
- `dlmoplotfine`: fine diagnostic plot, if `fine_flag = TRUE`

## Choosing a threshold

The `threshold` argument sets the melatonin concentration used to define the
rise. The appropriate value depends on the assay, study protocol, analysis plan,
and the participant's overall melatonin production; for example, a profile with
lower melatonin concentrations may require a lower threshold than a profile with
higher concentrations. Changing the threshold can change the base segment, the
ascending segment, and whether enough points remain to fit the hockey-stick
model.

Profile C illustrates this. At `threshold = 2.7`, the selected region does not
leave enough usable fitting points:

```{r}
profile_c <- system.file(
  "extdata/vignette_profiles/melatonin_profile_c.csv",
  package = "dlmoR"
)
```

The raw profile helps explain why. With this threshold, the line sits close to
the small early rise, so the selected ascending region can become too cropped to
support the hockey-stick fit:

```{r}
#| label: fig-profile-c-raw-threshold-27
#| fig-cap: "*Raw Profile C with threshold = 2.7. The threshold lies close to the small early rise, which can leave too little fitting context after segment selection.*"
plot_raw_profile(file_path = profile_c, threshold = 2.7)
```

```{r}
#| error: true
calculate_dlmo(
  file_path = profile_c,
  threshold = 2.7,
  fine_flag = FALSE
)
```

Raising the threshold to `4` better matches the intended rise for this profile:

```{r}
dlmo_profile_c_threshold_4 <- calculate_dlmo(
  file_path = profile_c,
  threshold = 4
)
```

```{r}
#| label: fig-threshold-profile-c
#| fig-cap: "*Profile C analysed with a higher threshold. Changing the threshold changes which points are eligible for the rise and can restore a usable fitting region.*"
dlmo_profile_c_threshold_4$dlmoplotfine
```

## Understanding interval_limit

Some melatonin profiles cross the threshold more than once. The `interval_limit`
argument controls how close repeated candidate rises are handled. It is useful
when an early above-threshold excursion is followed by a later, more plausible
melatonin rise.

The current repeated-crossing behavior is:

- a single point above threshold that immediately returns below threshold is
  treated as a blip and ignored automatically
- two or more consecutive points above threshold are treated as a candidate rise
- if another candidate rise starts within `interval_limit`, the later rise is
  taken to represent the melatonin rise
- if the later rise starts after `interval_limit`, the earlier candidate rise is
  kept

Numeric values are interpreted as hours, so `interval_limit = 2` is equivalent
to `lubridate::duration(hours = 2)`. Fractional values are allowed; for example,
`interval_limit = 0.5` is 30 minutes.

Importantly, `interval_limit` is measured between the first above-threshold
sample of each candidate rise. A below-threshold sample immediately before a
large increase may be used later by the hockey-stick fit, but it is not itself a
threshold crossing.

Profile B has an early candidate rise and a later stronger rise. With the
default 2-hour interval, the later rise starts too far away to replace the early
candidate:

```{r}
profile_b <- system.file(
  "extdata/vignette_profiles/melatonin_profile_b.csv",
  package = "dlmoR"
)
```

```{r}
dlmo_profile_b_default <- calculate_dlmo(
  file_path = profile_b,
  threshold = 2.7
)
```

```{r}
#| label: fig-profile-b-default
#| fig-cap: "*Profile B with the default interval limit. The earlier candidate rise is retained.*"
dlmo_profile_b_default$dlmoplotfine
```

Using a longer interval allows the later sustained rise to replace the early
candidate:

```{r}
dlmo_profile_b_interval <- calculate_dlmo(
  file_path = profile_b,
  threshold = 2.7,
  interval_limit = 8
)
```

```{r}
#| label: fig-profile-b-interval
#| fig-cap: "*Profile B with a longer interval limit. The later sustained rise is selected.*"
dlmo_profile_b_interval$dlmoplotfine
```

## When interval_limit is not needed

Not every repeated threshold crossing requires a wider interval limit. There
are two common cases where the default behavior may already be appropriate.

First, if there is only one point above threshold and the profile immediately
returns below threshold, `dlmoR` treats that point as a blip and ignores it
automatically.

Profile F shows this behavior. The small single-point excursion is not treated
as a candidate rise, so the later sustained rise is selected with the default
interval limit:

```{r}
profile_f <- system.file(
  "extdata/vignette_profiles/melatonin_profile_f.csv",
  package = "dlmoR"
)
```

```{r}
dlmo_profile_f <- calculate_dlmo(
  file_path = profile_f,
  threshold = 2.7
)
```

```{r}
#| label: fig-single-point-blip
#| fig-cap: "*Profile F contains a single-point above-threshold excursion. It is ignored automatically, so no wider interval limit is needed.*"
dlmo_profile_f$dlmoplotfine
```

Second, if a later candidate rise starts outside the `interval_limit`, it does
not replace the earlier selected rise. With the default setting, this means that
a later threshold crossing more than 2 hours after the earlier candidate is
ignored for repeated-crossing selection. In that case, the later crossing is not
missed because it is a blip; it is ignored because it is outside the interval
window.

Profile I shows the second case. Here, a later threshold crossing occurs after
the selected rise, but it starts well outside the default 2-hour interval
window. The later crossing is therefore ignored for repeated-crossing selection,
and the earlier candidate rise is retained:

```{r}
profile_i <- system.file(
  "extdata/vignette_profiles/melatonin_profile_i.csv",
  package = "dlmoR"
)
```

```{r}
dlmo_profile_i <- calculate_dlmo(
  file_path = profile_i,
  threshold = 2.7
)
```

```{r}
#| label: fig-outside-interval-window
#| fig-cap: "*Profile I contains a later threshold crossing, but it starts outside the default 2-hour interval window. The earlier candidate rise is retained.*"
dlmo_profile_i$dlmoplotfine
```

## Threshold and interval_limit can interact

Sometimes threshold and interval settings need to be considered together. A
threshold may work well for one profile but produce an early candidate rise in
another. A wider `interval_limit` can then be needed to tell the algorithm that
a later rise should replace the earlier candidate.

Profile H works with a higher threshold and the default interval:

```{r}
profile_h <- system.file(
  "extdata/vignette_profiles/melatonin_profile_h.csv",
  package = "dlmoR"
)
```

```{r}
dlmo_profile_h_threshold_5 <- calculate_dlmo(
  file_path = profile_h,
  threshold = 5
)
```

```{r}
#| label: fig-profile-h-threshold-5
#| fig-cap: "*Profile H with threshold = 5 and the default interval limit.*"
dlmo_profile_h_threshold_5$dlmoplotfine
```

At `threshold = 3`, the default interval limit is too short for the later rise
to replace the earlier candidate:

```{r}
#| error: true
calculate_dlmo(
  file_path = profile_h,
  threshold = 3,
  fine_flag = FALSE
)
```

In this profile, the early candidate rise starts at the first above-threshold
sample around 12:00. The later sustained rise visually begins after the
below-threshold point around 20:00, but its first above-threshold sample is
around 22:00. Therefore, the interval relevant to `interval_limit` is about 10
hours, not about 2 hours.

```{r}
dlmo_profile_h_threshold_3_interval <- calculate_dlmo(
  file_path = profile_h,
  threshold = 3,
  interval_limit = 10
)
```

```{r}
#| label: fig-profile-h-threshold-3-interval
#| fig-cap: "*Profile H with threshold = 3 and interval_limit = 10. The later sustained rise is selected because the interval is measured between candidate threshold-crossing starts.*"
dlmo_profile_h_threshold_3_interval$dlmoplotfine
```

## Slow rises and parallelogram truncation

After the ascending region is identified, `dlmoR` applies truncation rules so
that the rise used for the DLMO estimate is rapid and consistent. First, the
rightmost part of the ascending region is removed if the final segment has a
zero or negative slope.

The parallelogram rule is then applied to slow rises. An optimal parallelogram
is fitted around the ascending points and the last point before the ascending
region. The slope of the parallelogram's longest diagonal is compared with the
slope of its lateral edge. If the diagonal-to-lateral-edge slope ratio is not
greater than 1/2, the last segment of the ascending profile is removed. The
parallelogram is then recalculated, and the process is repeated until the rule
is satisfied.

Profile G showcases this truncation rule. After the initial rise, melatonin
production briefly lulls between roughly midnight and 03:00. That slower,
irregular portion is truncated so that the hockey-stick fit is based on the
rapid and consistent part of the rise. The plotted parallelogram shows the
ascending portion retained for the hockey-stick fit after the rule has been
applied:

```{r}
profile_g <- system.file(
  "extdata/vignette_profiles/melatonin_profile_g.csv",
  package = "dlmoR"
)
```

```{r}
dlmo_profile_g <- calculate_dlmo(
  file_path = profile_g,
  threshold = 2.7
)
```

```{r}
#| label: fig-slow-rise-parallelogram
#| fig-cap: "*Profile G illustrates a slow, irregular rise with a lull after midnight. The parallelogram rule truncates the ascending segment before fitting.*"
dlmo_profile_g$dlmoplotfine
```

## Explore the residual landscape

`calculate_dlmo()` searches over the region of interest containing possible
hockey-stick fits and chooses the solution with the lowest residual error. The
residual explorer is an
interactive Shiny app for inspecting that search space after a result has been
calculated.

```{r}
#| eval: false
run_residual_explorer(dlmo_result)
```

Use the residual explorer when you want to assess how stable or ambiguous a DLMO
estimate is. This is especially helpful when a profile has more than one
plausible rise, a noisy baseline, sparse sampling, or a broad region of interest.
In those cases, the best-fit DLMO may be technically valid but still worth
checking against nearby low-residual alternatives.

### How to read the explorer

The heatmap shows the residual error across the DLMO search space. The x-axis
is candidate DLMO time, the y-axis is candidate melatonin concentration at the
DLMO, and the colour scale shows the residual value for each candidate
hockey-stick fit. Darker colours indicate lower residuals. The white diamond
marks the DLMO selected by `calculate_dlmo()`.

The left-hand controls let you decide how much of the low-residual landscape to
summarize. The clustering threshold selects the lowest-residual grid points to
display, such as the best 1% or 5% of candidate fits. The minimum cluster size
sets how many neighbouring grid points are required before a group is treated as
a cluster rather than a tiny isolated region.

The violin plot summarizes the timing of the low-residual clusters. Each row is
a cluster of similarly good candidate DLMO solutions. The vertical reference
line marks the selected DLMO, and the horizontal spread shows how far each
cluster lies before or after that estimate. A tight cluster centered near zero
suggests a well-localized estimate; a broad cluster suggests a wider uncertainty
window; several separated clusters suggest competing candidate DLMO regions.

The summary tables quantify the same information. The nominal table gives the
size of each cluster and its mean, median, minimum, and maximum clock time. The
relative-time table expresses those cluster times as minutes before or after the
selected DLMO. These tables are useful when the heatmap suggests multiple
possible solutions and you want to know whether they differ by a few minutes or
by an hour or more.

### Multiple similarly good candidate regions

Profile D shows several low-residual clusters. The selected DLMO is 23:33:36.
With the explorer set to the lowest 1% of residuals, the residual values shown
are 0.027 to 0.027 across 263 grid points. In other words, at the printed
precision, these candidate solutions are essentially indistinguishable; any
differences are on the order of later decimal places.

```{r}
#| label: fig-residual-profile-d
#| echo: false
#| fig-cap: "*Residual explorer for profile D. Several separated clusters have nearly identical residual values, indicating multiple similarly good candidate DLMO regions.*"
knitr::include_graphics(
  article_figure("extdata/vignette_figures/residual_explorer_melatonin_profile_d.png")
)
```

The selected cluster is centered on the DLMO estimate, with a time range of
about -12.6 to +12.6 minutes relative to the selected DLMO. However, other
clusters with nearly identical residuals occur approximately 42.6 minutes
earlier, 84.0 minutes earlier, 43.2 minutes later, 85.8 minutes later, and
128.7 minutes later. This pattern means the chosen estimate is one of several
similarly good solutions. The diagnostic profile plot and the analysis settings
should be reviewed carefully before treating the selected time as uniquely
determined.

### One broad low-residual region

Profile E shows a different pattern. The selected DLMO is 20:55:36. With the
explorer set to the lowest 5% of residuals, the residual values shown are 0.278
to 0.278 across 572 grid points. Again, at the printed precision, many grid
points fit essentially equally well.

```{r}
#| label: fig-residual-profile-e
#| echo: false
#| fig-cap: "*Residual explorer for profile E. The low-residual region forms one broad cluster rather than several separated alternatives.*"
knitr::include_graphics(
  article_figure("extdata/vignette_figures/residual_explorer_melatonin_profile_e.png")
)
```

Unlike profile D, the low-residual solutions form one broad cluster rather than
several separated clusters. The cluster spans from about 0.6 minutes before the
selected DLMO to about 45.0 minutes after it, with a mean offset of 22.55
minutes. This suggests a broad continuous region of similarly good solutions,
rather than disjoint jumps between separate candidate DLMO regions. In practice,
that means the estimate is plausible, but the uncertainty window is broad. In
this example, that broad window is likely exacerbated by sparse melatonin
sampling across the relevant part of the rise, with no samples between roughly
22:00 and 02:00.

In practice, the residual explorer is most useful as a sensitivity check after
you have already looked at the diagnostic profile plot. The profile plot tells
you what the algorithm selected; the residual explorer tells you whether nearby
or competing solutions fit almost as well.

## Batch workflows

Because `calculate_dlmo()` returns ordinary R objects, `dlmoR` fits naturally
into scripted workflows. You can wrap the function in a small batch script to
process many melatonin profiles, save the full result objects for later
inspection, export diagnostic plots, and write a compact summary table with one
row per input file. The same idea works whether you iterate over file paths or
over data frames that have already been read into R.

The code below shows one example file-based workflow:

```{r}
#| eval: false
files <- list.files("path/to/profiles", pattern = "\\.csv$", full.names = TRUE)

output_dir <- "dlmo_batch_output"
plot_dir <- file.path(output_dir, "plots")
object_dir <- file.path(output_dir, "objects")

dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(object_dir, recursive = TRUE, showWarnings = FALSE)

dlmo_results <- lapply(files, function(file_path) {
  profile_name <- tools::file_path_sans_ext(basename(file_path))

  result <- calculate_dlmo(
    file_path = file_path,
    threshold = 5
  )

  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0(profile_name, "_coarse.png")),
    plot = result$dlmoplotcoarse,
    width = 8,
    height = 5,
    dpi = 300
  )

  ggplot2::ggsave(
    filename = file.path(plot_dir, paste0(profile_name, "_fine.png")),
    plot = result$dlmoplotfine,
    width = 8,
    height = 5,
    dpi = 300
  )

  saveRDS(
    result,
    file = file.path(object_dir, paste0(profile_name, "_dlmo_result.rds"))
  )

  result
})

names(dlmo_results) <- tools::file_path_sans_ext(basename(files))

summary_table <- data.frame(
  file_name = basename(files),
  profile_name = names(dlmo_results),
  threshold = vapply(dlmo_results, function(x) x$threshold, numeric(1)),
  interval_limit = vapply(dlmo_results, function(x) as.character(x$interval_limit), character(1)),
  dlmo_coarse = vapply(dlmo_results, function(x) as.character(x$dlmo$coarse$time), character(1)),
  dlmo_fine = vapply(dlmo_results, function(x) as.character(x$dlmo$fine$time), character(1))
)

writexl::write_xlsx(
  summary_table,
  path = file.path(output_dir, "dlmo_summary.xlsx")
)

saveRDS(
  dlmo_results,
  file = file.path(output_dir, "dlmo_results_all_profiles.rds")
)
```
